---
title: "CMIP6VisR"
author: "Simon Michael Papalexiou, Kevin Shook, Konstantinos Andreadis, Heba Abdelmoaty"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    number_sections: yes
vignette: >
  %\VignetteIndexEntry{CMIP6VisR}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
  header-includes:
   - \usepackage{amsmath}
editor_options: 
  chunk_output_type: inline
---

***

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = TRUE,
                      echo = TRUE,
                      fig.width = 7, 
                      warning = FALSE,
                      message = FALSE)
```
`CMIP6VisR` aims to facilitate handling the downscaled product of CMIP6 projectins **SPQM-CMIP6_CAN**. 

This dataset provides high-resolution (0.1°) daily precipitation and temperature projections for Canada, extending to 2100. Bias correction was performed using the Semi-Parametric Quantile Mapping (SPQM) method, ensuring reliable representation of observed marginal properties, trends, and variability while enabling a smooth transition from historical observations to future projections. This dataset includes 693 precipitation simulations from 34 climate models and 581 temperature simulations from 27 models, covering four Shared Socioeconomic Pathways (SSPs). Designed for hydrological, environmental, and ecological research, SPQM-CMIP6-CAN supports risk assessment, planning, and uncertainty quantification in climate impact studies.


<br>
~_Bias-corrected high-resolution temperature and precipitation projections for Canada_ | [Abdelmoaty et al. (2025)](https://www.nature.com/articles/s41597-025-04396-z?)

_"This publication offers the dataset along with an extensive validations on the daily scales. Our validation focuses on two key aspects: temporal and spatial validation. Temporally, we demonstrate the seamless continuity of the SPQM-CMIP6-CAN dataset with observations across all the shared socio-economic pathways (SSPs) of CMIP6, ensuring it preserves the observed distribution while capturing future trends in magnitude and non-stationarity. Spatially, we assess the effects of spatial aggregation on data integrity and visualize precipitation and temperature distributions through regional maps. The dataset remains available at a daily resolution, allowing users to conduct further region-specific validations."_

~_Changes in the risk of extreme temperatures in megacities worldwide_ | [Rajulapati et al. (2022)](https://doi.org/10.1016/j.crm.2022.100433)

_"This publication employed the same bias correction method for daily temperature to investigate future minimum and maximum temperature projections for 199 megacities worldwide. The changes in maximum and minimum temperature are quantified in terms of climate indices (ETCCDI and HDWI) for the four SSPs (SSP1-2.6, SSP2-4.5, SSP3-7.0, and SSP5-8.5)."_

~_Precipitation Bias Correction: A Novel Semi-parametric
Quantile Mapping Method_ | [Rajulapati and Papalexiou (2023)](https://doi.org/10.1029/2023EA002823)

_" This publication refines the bias correction method for daily precipitation by accounting for wet and dry spells is implemented in this publication, showing a case study in the Bow River Basin"_

~_Exacerbated heat in large Canadian cities_ | [Rajulapati et al. (2022)](https://doi.org/10.1016/j.uclim.2022.101097)


_"This publication employes the SPQM-CMIP6-CAN to explore future trends of minimum and maximum temperatures for the 18 most populated Canadian cities."_

~_Changes of Extreme Precipitation in CMIP6 Projections: Should We Use Stationary or Nonstationary Models?_ | [Abdelmoaty et al. (2023)](https://doi.org/10.1175/JCLI-D-22-0467.1)


_"This publication employes the same methods for bias correctiong daily precipitation to explore future extreme precipitation and their non-stationarity globally"_

***
# Step by step guide to facilitate employing SPQM-CMIP6-CAN {#section_1}

Install packages 
```{r, warning=FALSE, message=FALSE}
library(CMIP6VisR)
library(sf)
library(terra)
```


## Download the required simulation


Each simulation is divided into nine zones to facilitate efficient computational processing. As a first step, users must determine which zone(s) their shapefile intersects with to ensure they download the correct simulation data.


To know which zones are included in the targeted region, refer to the following chunk:


```{r warning=FALSE}
# convert the dataframe of zones to raster (built-in the package)
zone_area_grid <- cv_zone_area_raster() 


# Read the required shapefile
fpath <- system.file("extdata", "07BF001.shp", package = "CMIP6VisR")
basin_vector <- vect(fpath)


req_data <- cv_clip_basin(zone_area_grid, basin_vector)


#Printing the zones that are required for the targeted area
print(req_data[["zone"]])
```


```{r message=FALSE, warning=FALSE}
# Isolating the raster outputs
raster_list <- req_data$raster

# Merge the individual rasters using terra::merge
merged_raster <- do.call(terra::merge, req_data$raster)

# Plot the merged raster
terra::plot(merged_raster, main = expression("Grid Areas in " ~ km^2))


```





Once the relevant zone(s) are identified, users can proceed to download the required simulations, selecting the appropriate climate model and future scenario. These files can be accessed and downloaded locally via the FRDR portal: https://www.frdr-dfdr.ca/repo/dataset/7a02a1e0-3402-4183-af41-dfa483ef06b1

Alternatively, users have the option to leverage Globus Data Transfer for downloading files directly through the FRDR portal, providing a seamless transfer of large datasets. Users will need to set up their own Globus account, and thus this data could easily be transferred to local directories or HPCs. 



## Get the data within the required shapefile


